---
title: "SPECIAL_cohort"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SPECIAL_cohort}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=F
)
```
## SPatial transcriptomics cEll-Cell Interaction ALgorithm

**Attention**

The tutorial provided uses very small datasets to demonstrate how to use SPECIAL (+SOCIAL) on **Cohort** studies. Due to the very large memory required to analyze spatial transcriptomics, we require that users run SPECIAL on a HPC supercomputing cluster, such as the NIH Biowulf. This script will reproduce the SPECIAL results published by Sahni et al.

## 1. Installation

```{r install}
require(devtools)
devtools::install_github("sahil-sahni/5. Package(s)/SOCIAL")
```

## 2. Library

```{r library}
library(SOCIAL)
```

## 3. Organized inputs for Cohort analysis
SPECIAL inputs require specific formatting. These inputs are similar to SOCIAL, with the exception that location information is accounted for. When doing cohort analyses with multiple transcriptomics slides, we require the input to be organized in a specific named list format. In addition, these inputs must be saved as a RDS file before running SPECIAL. See GitHub folder 3. SPECIAL & SOCIAL for example code on how to generate these SPECIAL input files.

Example SPECIAL inputs include:

1. [biermann2022_SPECIAL_input.RDS](https://zenodo.org/records/13172848)
2. [thrane2018_SPECIAL_input.RDS](https://zenodo.org/records/13172848)

```{r cohort}

tumorslide1 = list(name = 'tumorslide1', # name of tumor slide sample (string)
                   pairs = pairs, # see other vignettes (matrix) 
                   loc = loc,  # see other vignettes (data.frame)
                   expr = expr, # see other vignettes (matrix)
                   ) #must be named list

SPECIAL_input = list(tumorslide1=tumorslide1, tumorslide2=tumorslide2) #must be named list

saveRDS(file='SPECIAL_input.RDS', SPECIAL_input)
```

## 4. SPECIAL for cohort studies
### Step 1 & 2: Divide spatial slides into regions and run SOCIAL step 1 through 3 on each region 

We require users to run SPECIAL step 1 & 2 on a HPC supercomputing cluster to harness parallelization with rslurm.

#### Biermann et al. 2022

```{r}
## ------- Input -------
# load inputs
path_input = 'biermann2022_SPECIAL_input.RDS'

## ------- Initialize environment -------
## set seed 
set.seed(20892) #NIH zip code

## ------- Run SPECIAL -------
n_iterations = 100
distance = 250
platform = 'slideseqv2'
puck_size = 3000

output = SPECIAL.cohort(path_input, n_iterations, distance, platform, puck_size)

saveRDS(output, file='biermann22_SPECIAL_output.RDS')
```

#### Thrane et al. 2018

```{r}
## ------- Input -------
# load inputs
path_input = 'thrane2018_SPECIAL_input.RDS'

## ------- Initialize environment -------
## set seed 
set.seed(20892) #NIH zip code

## ------- Run SPECIAL -------
n_iterations = 100
distance = 1
platform = 'BULK'

output = SPECIAL.cohort(path_special, path_input, n_iterations, distance, platform)

saveRDS(output, file='thrane2018_SPECIAL_output.RDS')
```
