---
title: "IRIS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IRIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Immunotherapy Resistance cell-cell Interaction Scanner

This tutorial will teach you how to use IRIS on deconvolved bulk transcriptomics data (via CODEFACS & LIRICS). Due to the computational intensiveness of this algorithm, we highly recommend running the following code on a HPC super computer cluster, like Biowulf. 

For this tutorial we recommend that you download the following input data:
1. [IRIS_LIRICS_final_ICB_cohort_input_response.Rdata](https://zenodo.org/records/13172848)

## Installation

```{r install}
require(devtools)
devtools::install_github("sahil-sahni/5. Package(s)/IRIS")
```

## Load library

```{r library}
library(devtools)
```

## Inputs 

Please download sample input files here: XX. 
Design IRIS specific inputs following the example data shown in **```IRIS_LIRICS_final_ICB_cohort_input_response.Rdata```**.

For each cohort, inputs should be organized as shown below, with **`sample`** and **`response`** columns. User can have as many cell-cell interaction columns where the values are the CCIs binary activity (inferred via [**LIRICS**](https://pubmed.ncbi.nlm.nih.gov/34983745/)) in a given tumor/patient sample. Combine multiple cohort inputs as lists for IRIS and **specify** if the cohort is of pre or post/on treatment samples (do **NOT** mix).

note 1: replace XX with cohort name (do not use "_" in the middle of the name)


```{r input}
## cohort input structure
cohort_input = data.frame(sample=c('A','B','C','D'), # patient/tumor sample barcode (no string specific format)
                        response=c(0,1,0,1), # patient response (1 or Responder; 0 for Non-Responder)
                        LigandCell_ReceptorCell_Ligand_Receptor=c(0,1,0)) # cell-cell interaction (1 for active; 0 for inactive based on LIRICS)*
   
## IRIS input
IRIS_input = list(Gide_pre=cohort_input, Gide_post=cohort_input, Riaz_post=cohort_input, Liu_pre=cohort_input)

```
*LigandCell_ReceptorCell_Ligand_Receptor* is the preferred name (output of LIRICS), however IRIS can take as input any string format.

## Run IRIS

a. to extract features using Leave One Out Cohort Validation (LOOCV)
b. to extract features directly from IRIS
c. to extract features from IRIS (alternative)

### a.) to extract features using Leave One Out Cohort Validation 

This method (used in paper) streamlines the ability to indendently validate the features you extract from IRIS onto numerous external cohorts. This works best when you have numerous pre- and post/on- treatment cohorts and you want to iteratively evaluate IRIS on each cohort independently.  

**1. inputs**

 specify validation/testing cohort (features to be applied to) and step 1&2 training cohort (features to be learned from)
 
  - test/validation (**`test_input`**) list must include at least one cohort (XX_pre or XX_on)
  - step 1 (**`s1_input`**) list must include at least one pre (XX_pre) and one post/on (XX_on) treatment cohort 
  - step 2 (**`s2_input`**) list must include at least one pre treatment cohort (XX_pre)

note 2: it is worth noting that the test/validation cohort (XX_pre or XX_on) does not have to be found in either the step 1 and step 2 cohort list. If this is the case, all the cohorts in step 1 and step 2 will be used for training.

```{r 1.1}
## additional function 
loadRData <- function(fileName){
  #loads an RData file, and returns it
  load(fileName)
  get(ls()[ls() != "fileName"])
}

## load IRIS inputs
IRIS_input = loadRData("IRIS_LIRICS_final_ICB_cohort_input_response.Rdata")
IRIS_input = IRIS_input[[1]]

test_input = IRIS_input #test input(s) is a list!
s1_input = IRIS_input #step 1 input(s) is a list!
s2_input = IRIS_input[c(2,3,5,8)] #step 2 input(s) is a list!
```

**3. specify direction and run LOOCohort Validation**

Extract IRIS features using LOOCohort Validation. IRIS only outputs **features** either RDI or RUI that can be scored to be applied to the approrpiate testing cohort ("testing cohort name").
```{r loocv}
## Perform LOOCohort Validation with IRIS ~ extract features only
#specify direction
direction="RDI" #RDI for resistance downregulated; RUI for Resistance Upregulated

#run LOOC_validation
output=lapply(1:length(test_input), function(e) IRIS::LOOC_validation(
  test_name = names(test_input[e]),
  s1_cohort = s1_input, 
  s2_cohort = s2_input,
  direction=direction))

names(output) = names(test_input)
```

**4. IRIS output**

When performing IRIS, the output is as follows. List index:

1. **testing cohort name**: the cohort to which you should score your IRIS features with.
2. **step 2 pre-treatment cohort**: the pre-treatment cohort that was utilized for step 2 feature selection.

```{r loocv output}
## IRIS output structure
output[["testing cohort name"]][["step 2 pre-treatment cohort"]]

# example
output[["Gide_pre"]] # extract features that will be used to score/test on Gide_pre
output[["Gide_pre"]][["Riaz_pre"]] # extract features that were inferred using Riaz_pre as step 2 training cohort ONLY and will be used to score/test on Gide_pre
```

**5. (optional) score IRIS interactions using ensembl model**

To replicate scoring using an ensemble model as in Sahni et al. please run the code below. Alternative scoring methods are possible and encouraged depending on the number of training cohorts available etc. and question being asked.

```{r score}

## load scoring input
test_input = IRIS_input # cohort to be scored/tested on (is a list!)
test_name = names(test_input) # name of testing cohort
features = output # features to be used on testing cohort (see step 4 example)
sample_name = T # boolean to specify if sample name and test cohort name should appear in output

cohort_score=lapply(1:length(test_input), function(or) calculate_score(
  test = test_input[[or]], 
  test_name = test_name[[or]], 
  features = features[[or]],
  sample_name = sample_name))

cohort_score=do.call(rbind,cohort_score) 

```

### b.) to extract features directly from IRIS 

This method allows one to run IRIS directly without going through leave one out cohort validation. This method is ideal if the user wants to run IRIS once *OR* if the user wants to train on all the data available to them. It is important to note that IRIS lets you utilize as many cohorts for step 1, but ONLY allows you to use ONE step 2 cohort for training. 

**2. specify IRIS input**

- step 1 (`s1_cohort`) list must include at least one pre and one on/post-treatment ICB cohort(s) (XX_pre/on)
- step 2 (`s2_cohort`) data.frame must include **ONLY ONE** pre-treatment ICB cohort (XX_pre)
- step 2 cohort name (`s2_name`) vector must be written as "XX_pre" (*note 3*)

*note 3*: it is important to specify step 2 cohort name (s2_name) in the format (XX_pre) and is consistent as a input for step 1 input and step 2 input. The goal of s2_name is to prevent overfitting by removing the step 2 cohort (both XX_pre and XX_on) from the first step feature selection. For example if you had a list of cohorts (i.e. Gide_pre, Riaz_pre, Gide_on, Riaz_on, and Liu_pre), you specified your s1_cohorts to be (Gide_pre, Riaz_pre, Gide_on, Riaz_on, and Liu_pre) and your s2_cohorts to be (Gide_pre), and you specified your step 2 cohort name as ('Gide_pre'), then IRIS will automatically remove Gide_pre and Gide_on from the first step training and will only utilize Gide_pre in the second step training. Note that if there is no step 1 cohorts (s1_cohort) available for IRIS to train on, and error will occur. See (**c**) for alternative usage of IRIS for when you are limited with the training data available and want to override the mutual exclusive training argument

```{r b}

## load IRIS inputs
IRIS_input = loadRData("IRIS_LIRICS_final_ICB_cohort_input_response.Rdata")
IRIS_input = IRIS_input[[1]]

s1_input = IRIS_input #step 1 input(s) is a list!
s2_input = IRIS_input[[c(2)]] #step 2 input (ex. Gide_pre) is a data.frame!
s2_name = 'Gide_pre' #step 2 name is a string!

direction = 'RDI' #direction is a string
```

**3. run IRIS and extract features**
```{r IRIS}
## run IRIS
output = IRIS::IRIS(s2_input, s2_name, s1_input, direction) #output is a single vector of RDIs or RUIs inferred
```

### c.) to extract features directly from IRIS (alternative)

This method is ideal for when you are limited in the number of training cohorts available to you OR if you would like to reutilize the cohort from step 1 in step 2 feature selection. 

**2. specify IRIS input**

- step 1 (`s1_cohort`) list must include at least one pre and one on/post-treatment ICB cohort(s) (XX_pre/on). 
- step 2 (`s2_cohort`) data.frame must include ONLY ONE pre-treatment ICB cohort (XX_pre)
- step 2 cohort name (`s2_name`) vector must be written as "NULL_pre" (*note 4*)

```{r c}

## load IRIS inputs
IRIS_input = loadRData("IRIS_LIRICS_final_ICB_cohort_input_response.Rdata")
IRIS_input = IRIS_input[[1]]

s1_input = IRIS_input #step 1 input(s) is a list!
s2_input = IRIS_input[[c(2)]] #step 2 input (ex. Gide_pre) is a data.frame!
s2_name = 'NULL_pre' #step 2 name is a string!

direction = 'RDI' #direction is a string
```

*note 4*: it is important to specify step 2 cohort name (s2_name) as ('NULL_pre') to reutilize a cohort for both step 1 and step 2. You do not need to rename step 1 or step 2 cohorts, IRIS will automatically use step 2 cohort in step 1 when specifying the name as "NULL_pre". For example if you had a list of cohorts (i.e. Gide_pre, Riaz_pre, Gide_on, Riaz_on, and Liu_pre), you specified your s1_cohorts to be (Gide_pre, Riaz_pre, Gide_on, Riaz_on, and Liu_pre) and your s2_cohorts to be (Gide_pre), and you specified your step 2 cohort name as ('NULL_pre'), then IRIS will not remove Gide cohort from step 1. Make sure that NULL_pre is not a name of a cohort!

**3. run IRIS and extract features**

```{r IRIS alternative}
## run IRIS
output = IRIS::IRIS(s2_input, s2_name, s1_input, direction) #output is a single vector of RDIs or RUIs inferred
```


