---
title: "IRIS_NatComms_Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IRIS_NatComms_Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

**1. load IRIS**

```{r setup}
devtools::install_github("sahil-sahni/5. Package(s)/IRIS")
library(IRIS)
```

**2. load additional functions**

these functions are not apart of the IRIS package
```{r function}
## ------- Additional Functions -------

# to load Rdata
loadRData <- function(fileName){
  #loads an RData file, and returns it
  load(fileName)
  get(ls()[ls() != "fileName"])
}
```

**3. initialize parallelization environment**

```{r intialize}
## parallelize code
ncpus <- parallelly::availableCores()
print(ncpus)
options(mc.cores = ncpus) # set a global option for parallel packages
```

**4. load training inputs**

Download data here:
[IRIS_LIRICS_final_ICB_cohort_input_response.Rdata](https://zenodo.org/records/13172848)


```{r input}
## training input
model_input = loadRData("IRIS_LIRICS_final_ICB_cohort_input_response.Rdata") 
```

**5. specify IRIS inputs 

step 1 cohorts (pre + on), step 2 cohorts (pre only), and validation/testing cohort

```{r specify}
test_input = model_input[[1]] #test input(s)
s1_input = model_input[[1]] #step 1 input(s)
s2_input = model_input[[1]][c(2,3,5,8)] #step 2 input(s) - pre-treatment only
```

**6. specify direction (RDI or RUI)**

```{r direction}
direction = 'RDI' #resistance downregulated interactions
```

**7. perform leave one-out cohort validation with IRIS**

this function extracts features ONLY

```{r LOOCV}
feature_RDI=lapply(1:length(test_input), function(e) IRIS::LOOC_validation(names(test_input[e]), s1_input, s2_input, direction=direction))
names(feature_RDI) = names(test_input)

```

**8. score interactions in testing cohort with ensemble model**

using ensemble scoring as in Sahni et al. 2024.
```{r score}
cohort_score=lapply(1:length(test_input), function(or) calculate_score(test_input[[or]], names(test_input[or]), feature_RDI[[or]]))
cohort_score=do.call(rbind,cohort_score)
```
